---
title: "*Spatial Statistics Lab 5*"
author: '**Onyedikachi J Okeke**'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **0.0 To Load the library**
```{r,comment="",echo=TRUE,eval=FALSE}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)

```
##### Library 
```{r,echo=F,results='Show', fig.show='hold',eval=T}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
```
### **Q1 Loading Sids data**
```{r,comment="",echo=TRUE,eval=FALSE}
# Define the file path
file_path <- "C:/Spatial Statistics Labwork/Lab5Data/sids2.shp"

# Import the shapefile
sids <- st_read(file_path)

class(sids)

# Check the structure of the imported object
str(sids)

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Define the file path
file_path <- "C:/Spatial Statistics Labwork/Lab5Data/sids2.shp"

# Import the shapefile
sids <- st_read(file_path)

class(sids)

# Check the structure of the imported object
str(sids)

```

### **Check the projection of the shapefile**
```{r,comment="",echo=TRUE,eval=FALSE}
# Extract CRS information using st_crs()
crs_info <- st_crs(sids)

# Print CRS information
print(crs_info)

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Extract CRS information using st_crs()
crs_info <- st_crs(sids)

# Print CRS information
print(crs_info)

```

### **Define the target CRS (NAD27 UTM Zone 17N**
```{r,comment="",echo=TRUE,eval=FALSE}

# Assign the coordinate system (WGS84)
st_crs(sids) <- st_crs("+proj=longlat +ellps=WGS84")

# Check the current CRS
print(st_crs(sids))

# Define the target CRS (NAD27 UTM Zone 17N)
target_crs <- "+proj=utm +zone=17 +datum=NAD27"

# Project the shapefile to the target CRS
sids_projected <- st_transform(sids, target_crs)

# Check the CRS of the projected shapefile
print(st_crs(sids_projected))

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}

# Assign the coordinate system (WGS84)
st_crs(sids) <- st_crs("+proj=longlat +ellps=WGS84")

# Check the current CRS
print(st_crs(sids))

# Define the target CRS (NAD27 UTM Zone 17N)
target_crs <- "+proj=utm +zone=17 +datum=NAD27"

# Project the shapefile to the target CRS
sids_NAD <- st_transform(sids, target_crs)

# Check the CRS of the projected shapefile
print(st_crs(sids_NAD))

```

### **Define the CRS for North Carolina NAD 83 State Plane**
```{r,comment="",echo=TRUE,eval=FALSE}
# Define the CRS for North Carolina NAD 83 State Plane
nc_crs <- st_crs("+proj=lcc +lat_1=34.33333333333334 +lat_2=36.16666666666666 +lat_0=33.75 +lon_0=-79 +x_0=609601.22 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")

# Transform the sids shapefile to North Carolina NAD 83 State Plane
sids_SP <- st_transform(sids, nc_crs)
print(st_crs(sids_SP))

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Define the CRS for North Carolina NAD 83 State Plane
nc_crs <- st_crs("+proj=lcc +lat_1=34.33333333333334 +lat_2=36.16666666666666 +lat_0=33.75 +lon_0=-79 +x_0=609601.22 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")

# Transform the sids shapefile to North Carolina NAD 83 State Plane
sids_SP <- st_transform(sids, nc_crs)
print(st_crs(sids_SP))

```

### **Plotting the data in three different coordinate systems**
```{r,comment="",echo=TRUE,eval=FALSE}
par(mfrow=c(2,2))

plot(sids, axes=T)
title("WGS84")

plot(sids_NAD, axes = TRUE)
title("NAD27")

plot(sids_SP, axes=T)
title("NC State Plane")

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
par(mfrow=c(2,2))

plot(sids, axes=T)
title("WGS84")

plot(sids_NAD, axes = TRUE)
title("NAD27")

plot(sids_SP, axes=T)
title("NC State Plane")

```

### **Contiguity based neighbors**
```{r,comment="",echo=TRUE,eval=FALSE}
# Create Queen contiguity spatial weights matrix
sids_nbq <- poly2nb(sids, queen = TRUE)
# Check the summary of the spatial weights matrix
summary(sids_nbq)

# Create Queen (false) contiguity spatial weights matrix
sids_nbr <- poly2nb(sids, queen = FALSE)
# Check the summary of the spatial weights matrix
summary(sids_nbr)

plot(st_geometry(sids), main="Spatial Plot with Neighborhoods")

coords <- st_coordinates(sids)

# Check the number of neighborhoods in sids_nbq
num_neighborhoods <- length(sids_nbq)

# Check the number of coordinate pairs in coords
num_coords <- nrow(coords)

# Print out the counts for verification
print(paste("Number of neighborhoods:", num_neighborhoods))
print(paste("Number of coordinate pairs:", num_coords))

# Example adjustment (assuming you need to align coords_2d with sids_nbq correctly)
coords_2d_corrected <- coords[1:num_neighborhoods, ]

par(mfrow=c(1,1))

plot(sids_nbq, coords_2d_corrected)
plot(sids, add= TRUE)

plot(sids_nbr, coords_2d_corrected)
plot(sids, add= TRUE)

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Create Queen contiguity spatial weights matrix
sids_nbq <- poly2nb(sids, queen = TRUE)
# Check the summary of the spatial weights matrix
summary(sids_nbq)

# Create Queen (false) contiguity spatial weights matrix
sids_nbr <- poly2nb(sids, queen = FALSE)
# Check the summary of the spatial weights matrix
summary(sids_nbr)

plot(st_geometry(sids), main="Spatial Plot with Neighborhoods")

coords <- st_coordinates(sids)

# Check the number of neighborhoods in sids_nbq
num_neighborhoods <- length(sids_nbq)

# Check the number of coordinate pairs in coords
num_coords <- nrow(coords)

# Print out the counts for verification
print(paste("Number of neighborhoods:", num_neighborhoods))
print(paste("Number of coordinate pairs:", num_coords))

# Example adjustment (assuming you need to align coords_2d with sids_nbq correctly)
coords_2d_corrected <- coords[1:num_neighborhoods, ]

par(mfrow=c(1,1))

plot(sids_nbq, coords_2d_corrected)
plot(sids, add= TRUE)

plot(sids_nbr, coords_2d_corrected)
plot(sids, add= TRUE)


```
Q2

### **Q3 Plot the ROOK case**
```{r,comment="",echo=TRUE,eval=FALSE}
# Get coordinates from spatial data
coords <- st_coordinates(sids_SP) 

# Assuming the first two columns are latitude and longitude
coords_2d <- coords[, 1:2]

# Then, re-run the knn2nb process with the adjusted 2D coordinates
sids_kn1 <- knn2nb(knearneigh(coords_2d, k=1))
sids_kn2 <- knn2nb(knearneigh(coords_2d, k=2))
sids_kn4 <- knn2nb(knearneigh(coords_2d, k=4))

# Plot
plot(sids_SP, main = "SIDS Spatial Points")
plot(sids_kn1, coords, add = TRUE, col = "red")
plot(sids_kn2, coords, add = TRUE, col = "blue")
plot(sids_kn4, coords, add = TRUE, col = "green")

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Get coordinates from spatial data
coords <- st_coordinates(sids_SP) 

# Assuming the first two columns are latitude and longitude
coords_2d <- coords[, 1:2]

# Then, re-run the knn2nb process with the adjusted 2D coordinates
sids_kn1 <- knn2nb(knearneigh(coords_2d, k=1))
sids_kn2 <- knn2nb(knearneigh(coords_2d, k=2))
sids_kn4 <- knn2nb(knearneigh(coords_2d, k=4))

# Plot
plot(sids_SP, main = "SIDS Spatial Points")
plot(sids_kn1, coords, add = TRUE, col = "red")
plot(sids_kn2, coords, add = TRUE, col = "blue")
plot(sids_kn4, coords, add = TRUE, col = "green")


```

Q4.briefly describe their differences

### **Assign neighbors based on a specified distance**
```{r,comment="",echo=TRUE,eval=FALSE}
# Calculate distances
dist <- unlist(nbdists(sids_kn1, coords_2d))
summary(dist)  # Summarize distances to understand distribution
max_k1 <- max(dist)  # Find the maximum distance

# Create distance-based neighborhoods
sids_kd1 <- dnearneigh(coords_2d, d1=0, d2=0.75*max_k1)
sids_kd2 <- dnearneigh(coords_2d, d1=0, d2=1*max_k1)
sids_kd3 <- dnearneigh(coords_2d, d1=0, d2=1.5*max_k1)

# Set up the plotting area to have 3 rows for the different distance thresholds
par(mfrow=c(3,1))

# Plot for dist=0.75*max_k1
plot(st_geometry(sids_SP), main="dist=0.75*max_k1")  # Plot the spatial object
plot(sids_kd1, coords, add=TRUE)  # Add the dnearneigh plot

# Plot for dist=1*max_k1
plot(st_geometry(sids_SP), main="dist=1*max_k1")  # Plot the spatial object again for a new comparison
plot(sids_kd2, coords, add=TRUE)  # Add the dnearneigh plot

# Plot for dist=1.5*max_k1
plot(st_geometry(sids_SP), main="dist=1.5*max_k1")  # Plot the spatial object again for another new comparison
plot(sids_kd3, coords, add=TRUE)  # Add the dnearneigh plot

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Calculate distances
dist <- unlist(nbdists(sids_kn1, coords_2d))
summary(dist)  # Summarize distances to understand distribution
max_k1 <- max(dist)  # Find the maximum distance

# Create distance-based neighborhoods
sids_kd1 <- dnearneigh(coords_2d, d1=0, d2=0.75*max_k1)
sids_kd2 <- dnearneigh(coords_2d, d1=0, d2=1*max_k1)
sids_kd3 <- dnearneigh(coords_2d, d1=0, d2=1.5*max_k1)

# Set up the plotting area to have 3 rows for the different distance thresholds
par(mfrow=c(1,1))

# Plot for dist=0.75*max_k1
plot(st_geometry(sids_SP), main="dist=0.75*max_k1")  # Plot the spatial object
plot(sids_kd1, coords, add=TRUE)  # Add the dnearneigh plot

# Plot for dist=1*max_k1
plot(st_geometry(sids_SP), main="dist=1*max_k1")  # Plot the spatial object again for a new comparison
plot(sids_kd2, coords, add=TRUE)  # Add the dnearneigh plot

# Plot for dist=1.5*max_k1
plot(st_geometry(sids_SP), main="dist=1.5*max_k1")  # Plot the spatial object again for another new comparison
plot(sids_kd3, coords, add=TRUE)  # Add the dnearneigh plot

```
Q5. briefly describe their differences

### **Row-standardized weights matrix**
```{r,comment="",echo=TRUE,eval=FALSE}
#Row-standardized weights matrix
sids_nbq_w<- nb2listw(sids_nbq)
print(sids_nbq_w)
sids_nbq_w$neighbours[1:5]

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
#Row-standardized weights matrix
sids_nbq_w<- nb2listw(sids_nbq)
print(sids_nbq_w)
sids_nbq_w$neighbours[1:5]

```

Q6. Define neighbors based on a specified distance (the mean distance of neighbors). Take a screen shot of your plot and briefly describe their differences

### **Sids_nbq weights matrixs**
```{r,comment="",echo=TRUE,eval=FALSE}
#Row-standardized weights matrix
sids_nbq_w$weights[1:5]
```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
#Row-standardized weights matrix
sids_nbq_w$weights[1:5]
```

Q7. What is the result? What does this mean?


Q8. What is the result? What does this mean?

### **Binary Weights**
```{r,comment="",echo=TRUE,eval=FALSE}
sids_nbq_wb<-nb2listw(sids_nbq, style="B")
print(sids_nbq_wb)
sids_nbq_wb$weights[1:5]
```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
sids_nbq_wb<-nb2listw(sids_nbq, style="B")
print(sids_nbq_wb)
sids_nbq_wb$weights[1:5]
```

Q9. What is the result? What does this mean?

### **Inverse distance weighting bases weights on the distance between centroids**
```{r,comment="",echo=TRUE,eval=FALSE}
# Extract coordinates from the 'sf' object

coords_2d <- st_coordinates(sids_SP)[,1:2]

sids_nbq <- knn2nb(knearneigh(coords_2d, k=2)) 
dist <- nbdists(sids_nbq, coords_2d)

# Adjusting IDW calculation to handle distances of 0
idw <- lapply(dist, function(x) ifelse(x == 0, Inf, 1/(x/1000)))

# Creating a spatial weights object with the IDW values
sids_nbq_idwb <- nb2listw(sids_nbq, glist=idw, style="B")

# Accessing initial weights and providing summary
initial_weights <- sids_nbq_idwb$weights[1:5] # Get the first 5 weights
print(initial_weights)

# Summary of all weights
weights_summary <- summary(unlist(sids_nbq_idwb$weights))
print(weights_summary)

```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Extract coordinates from the 'sf' object

coords_2d <- st_coordinates(sids_SP)[,1:2]

sids_nbq <- knn2nb(knearneigh(coords_2d, k=2)) 
dist <- nbdists(sids_nbq, coords_2d)

# Adjusting IDW calculation to handle distances of 0
idw <- lapply(dist, function(x) ifelse(x == 0, Inf, 1/(x/1000)))

# Creating a spatial weights object with the IDW values
sids_nbq_idwb <- nb2listw(sids_nbq, glist=idw, style="B")

# Accessing initial weights and providing summary
initial_weights <- sids_nbq_idwb$weights[1:5] # Get the first 5 weights
print(initial_weights)

# Summary of all weights
weights_summary <- summary(unlist(sids_nbq_idwb$weights))
print(weights_summary)
```

Q10. What is the result? What does this mean?

### **Sids_nbq weights matrixs**
```{r,comment="",echo=TRUE,eval=FALSE}
# Summary of all weights
weights_summary <- summary(unlist(sids_nbq_idwb$weights))
print(weights_summary)
sids_nbq_w <- nb2listw(sids_nbq, zero.policy = T)
print(sids_nbq_w)
```
##### Result
```{r,echo=F,results='Show', fig.show='hold',eval=T}
# Summary of all weights
weights_summary <- summary(unlist(sids_nbq_idwb$weights))
print(weights_summary)
sids_nbq_w <- nb2listw(sids_nbq, zero.policy = T)
print(sids_nbq_w)
```
